---
title: "LRCoDa-vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LRCoDa-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## 1. Load Dependencies

```{r setup, message=FALSE, warning=FALSE}
library(LRCoDa)
library(robCompositions)
library(dplyr)
library(sjmisc)
library(robustbase)
library(forcats)
```

## 2. Load data

Load the dataset "gemas" from the package `robCompositions` and recode one level from "" to "Not Specified", because the function `LRCoDa` does not allow empty names for the levels.

```{r}
data("gemas")

gemas <- gemas %>% 
  mutate(soilclass = fct_recode(soilclass, "Not specified" = ""))
```

## 3. Introduction of the use of the factor variable extension with LRCoDa

With the newly released package `LRCoDa` we are able to use factor variables for linear regression on raw compositional data. As with the function `lmCoDaX` from the package `robCompositions` we have the choice between the "robust" and "classical" method.

First, we show an example where we apply the robust method to the dataset "gemas" and use the "soilclass" as the factor variable, the trace elements as composition and "sand" as target variable.

```{r, message=FALSE}
X <- dplyr::select(gemas, c(soilclass, Al:Zr))
y <- gemas$sand
LR_robust_1 <- LRCoDa(y = y, X = X, factor_column = 'soilclass', method = 'robust', 
                      max_refinement_steps = 500)
LR_robust_1$ilr
```

We can access the summary output of the Linear Regression as known from the function `lmCoDaX` with \$ilr (for the ilr output) and \$lm (for the lm output). In addition, we can clearly see which is the factor variable and which are the variables from the composition by the prefix of the variables. 

The output of the robust method is further differentiated from the output using `lmCoDaX` because we use `lmrob` for the robust calculation of the linear regression in `LRCoDa` instead of `ltsReg` in `lmCoDaX`.

### 3.1. Limitations on the use of factor variables

We list some scenarios where we show the limitations of using factor variables with the `LRCoDa` function.

#### 3.1.1. More than one factor variable is not allowed

Only can one factor variable can be used for the model. We return an error message if more than one factor variable is used.

```{r, error=TRUE, out.width=60}
X <- dplyr::select(gemas, c(soilclass, COUNTRY, Al:Zr))
LRCoDa(y = gemas$sand, X = X, factor_column = c('COUNTRY', 'soilclass'))
```

#### 3.1.2. Only characters and factors are allowed as datatypes for the factor variable

We only allow characters and factors as datatypes for the factor variable for the model. We return an error message if the factor variable contains a numeric value.

```{r, error=TRUE, out.width=60}
X <- dplyr::select(gemas, c(MeanTemp, Al:Zr))
LRCoDa(y = gemas$sand, X = X, factor_column = c('MeanTemp'))
```

#### 3.1.3. The factor has to be specified

When the factor column is not specified in the function input, but X contains a variable with factors or characters, the function returns an error message that "X contains non-numerical variables but there is no factor_column defined".

```{r, error=TRUE, out.width=60}
X <- dplyr::select(gemas, c(soilclass, Al:Zr))
LRCoDa(y = gemas$sand, X = X)
```

#### 3.1.4. Factor column must be specified as a name in the function input

The factor column must be specified exactly as in the help function described. It is not possible to fill in the whole column.

```{r, error=TRUE, out.width=60}
X <- dplyr::select(gemas, c(soilclass, Al:Zr))
LRCoDa(y = gemas$sand, X = X, factor_column = X %>% select(soilclass))
```

## 4. Introduction of the use of the external variable extension with LRCoDa

With the newly released package `LRCoDa` we are also able to use external variables that are not part of the composition. Here we also have the possibility to choose between the "robust" and "classical" method and can combine them with a factor variable.

We show an example where we combine external variables with a factor variables and of course the composition (naming as internal variables). We use the "gemas" dataset, where "MeanTemp" the external variable is, "soilclass" the factor variable, the trace elements the composition ("Internal" variables) and "sand" the target variable.

```{r, message=FALSE}
X <- dplyr::select(gemas, c(MeanTemp, soilclass, Al:Zr))
y <- gemas$sand
LR_robust_2 <- LRCoDa(y = y, X = X, external = c('MeanTemp'), factor_column = 'soilclass', 
                      method = 'robust', max_refinement_steps = 500)
LR_robust_2$ilr
```

We can see here that the different types of variables have different prefixes with which we can easily identify them.

### 4.1 Possibilities and Limitations of using external variables

We list the possibilities and limitations of using the external variables

#### 4.1.1. Multiple external variables are allowed and supported

We allow and support the use of multiple external variables in the same model.

```{r, error=TRUE, message=FALSE, out.width=60}
X <- dplyr::select(gemas, c(MeanTemp, AnnPrec, Al:Zr))
LRCoDa(y = gemas$sand, X = X, external = c('MeanTemp', 'AnnPrec'), 
       method = "classical")
```

#### 4.1.2. Only numerical variables are allowed as External variables

Only numeric variables are allowed for the external variables, as non-numeric variables are captured with the factor variables. We return an error message that says: "X contains non-numerical variables but there is no factor_column defined".

```{r, error=TRUE, out.width=60}
X <- dplyr::select(gemas, c(MeanTemp, AnnPrec, COUNTRY, Al:Zr))
LRCoDa(y = gemas$sand, X = X, external = c('MeanTemp', 'AnnPrec', 'COUNTRY'), 
       method = 'classical')
```

#### 4.1.3. External column has to be specified as name in the function input

The external column must be specified exactly as in the help function described. It is not possible to fill in the whole column.

```{r, error=TRUE, out.width=60}
X <- dplyr::select(gemas, c(MeanTemp, AnnPrec, Al:Zr))
LRCoDa(y = gemas$sand, X = X, external = X %>% select(c('MeanTemp', 'AnnPrec')))
```

## 5. Use of lmrob for robust fitting

With the newly introduced use of factor variables for linear regression on raw compositional data, we need to change the fitting algorithm from the old `ltsReg` (used in `lmCoDaX`) to `lmrob` (used in `LRCoDa`). This change offers many advantages when working with factor variables and also with external variables compared to `ltsReg` which is very limited for this application.
Also, `lmrob` determines robustness weights between 0 and 1, while `ltsReg` only works with 0 and 1 for the robustness weigths. However, there are also new challenges introduced with `lmrob`, as shown in the example below. Also, the backwards compatibility of the function when fitting with the robust method is no longer given due to this change.

```{r, error=TRUE, message=FALSE, out.width=60}
X <- dplyr::select(gemas, c(MeanTemp, AnnPrec, soilclass, Al:Zr))
LRCoDa(y = gemas$sand, X = X, external = c('MeanTemp', 'AnnPrec'), 
       factor_column = c('soilclass'), max_refinement_steps = 200)
```

The algorithm does not converge with the default refinement steps 200 as defined in the function `lmrob`. Therefore, we implemented the parameter max_refinement_steps as input to the function `LRCoDa` to change the maximum steps in such cases.

```{r, error=TRUE, message=FALSE, out.width=60}
X <- dplyr::select(gemas, c(MeanTemp, AnnPrec, soilclass, Al:Zr))
LRCoDa(y = gemas$sand, X = X, external = c('MeanTemp', 'AnnPrec'), 
       factor_column = c('soilclass'), max_refinement_steps = 500)
```

And now with max_refinement_steps set to 500 the algorithm converges.

## 6. further improvements

The norm of the pivot coordinate calculation can now also be defined in the function input, in contrast to the function `lmCoDaX`. The default value is "orthonorm" as it is preset in the function `pivotCoord` and therefore also the one used in `lmCoDaX` function.

```{r, error=TRUE, message=FALSE, out.width=60}
X <- dplyr::select(gemas, c(Al:Zr))
LRCoDa(y = gemas$sand, X = X, pivot_norm = "orthogonal", method = "classical")
```
